R for Clinical Study Reports and Submission

### 2021-09-15 tlf-disposition.Rmd
fix typo for disposition table

# Table for Study Patients

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize accounting of all patients who entered the study, 
in the section 10.1, disposition of patients of a CSR.

@@

The first step is to read relevant datasets into R. 
For disposition table, all the required information is saved in the ADSL dataset.
We can use `haven` package to read the dataset.

@@

```{r}
adsl %>%
  select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
  head(4)
```
In the code below, we calculated patients in the analysis population.

```{r}
n_rand <- adsl %>%
  group_by(TRT01PN) %>%
  summarize(n = n()) %>%
  pivot_wider(
    names_from = TRT01PN,
    names_prefix = "n_",
    values_from = n
  ) %>%
  mutate(row = "Patients in population")
n_rand
```
In the code below, we calculate number and percentage of patients who discontinued the study.

Here we use `formatC()` to customize the numeric value to be 1 decimal with fixed width of 5.

@@

n_disc
```
In the code below, we calculate number and percentage of patients 
who completed/discontinued the study in different reasons.

@@

In the code below, we calculate number and percentage of patients 
who complete the study. We split `n_reason` because we need to customize 
the row order of the table.

@@

n_complete
```
In the code below, we calculate number and percentage of patients 
who discontinued the study in different reasons. 
Here we use `paste0("    ", row)` to add some leading space for individual reasons for study discontinuation.

@@

Now we combined individual rows into the whole table for reporting purpose. 
`tbl_disp` is used as input for `r2rtf` to create final report.

@@

We start to define the format of the output. We highlighted items that 
is not discussed in previous discussion.

`rtf_title` define table title. We can provide a vector for the title argument. 
Each value is a separate line. 
The format can also be controlled by providing a vector input in text format.

@@

>>>>

# Disposition {#disposition}

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), we need to summarize the accounting of all patients who entered the study, in Section 10.1, Disposition of Patients.
The disposition of patients reports the numbers of patients who were randomized, and who entered and completed each phase of the study, as well as the reasons for all post-randomization discontinuations, grouped by treatment and by major reason (lost to follow-up, adverse event, poor compliance, etc.).

@@

The first step is to read relevant datasets into R.
For a disposition table, all the required information is saved in the ADSL dataset.
We can use the `haven` package to read the dataset.

@@

```{r}
adsl %>%
  select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
  head(4)
```

In the code below, we calculated patients in the analysis population.

```{r}
n_rand <- adsl %>%
  group_by(TRT01PN) %>%
  summarize(n = n()) %>%
  pivot_wider(
    names_from = TRT01PN,
    names_prefix = "n_",
    values_from = n
  ) %>%
  mutate(row = "Patients in population")
n_rand
```

In the code below, we calculate the number and percentage of patients who discontinued the study.

Here we use the `formatC()` function to customize the numeric value to be 1 decimal with fixed width of 5.

@@

n_disc
```

In the code below, we calculate the number and percentage of patients who completed/discontinued the study for different reasons.

@@

In the code below, we calculate the number and percentage of patients who complete the study.
We split `n_reason` because we need to customize the row order of the table.

@@

n_complete
```

In the code below, we calculate the number and percentage of patients who discontinued the study for different reasons.
Here we use `paste0("    ", row)` to add some leading space for individual reasons for study discontinuation.

@@

Now we combined individual rows into the whole table for reporting purposes.
`tbl_disp` is used as input for `r2rtf` to create the final report.

@@

We start to define the format of the output.
We highlighted items that is not discussed in the previous discussion.

The `rtf_title` defines table title.
We can provide a vector for the title argument.
Each value is a separate line.
The format can also be controlled by providing a vector input in text format.

@@

The procedure to generate a disposition table can be summarized as follow:

-   Step 1: Read data into R, i.e. `adsl`.
-   Step 2: Count patients in the analysis population and name the dataset as `n_rand`.
-   Step 3: Count the number and percentage of patients who discontinued the study, and name the dataset as `n_disc`.
-   Step 4: Count the number and percentage of patients who discontinued the study in different reasons, and name the dataset as `n_reason`. Format it using `r2rtf`.
-   Step 5: Count the number and percentage of patients who completed the study, and name the dataset as `n_complete`.
-   Step 6: Bind `n_rand`, `n_disc`, `n_reason` and `n_complete` by row. Format it using `r2rtf`.
